Sequence assembly

### 2025-04-09 episodes/sequence-assembly.Rmd
Finished assembly & annotation

@@

Hint:

@@

prints the 'help'

@@

Find out how many contigs there are in the *M. tuberculosis* isolates.

@@

@@

@@

@@

@@

```bash
for accession in ERR029207 ERR029206 ERR026478 ERR026474 ERR026473 ERR026481 ERR026482
do
  echo "${accession}"
  gt seqstat "${accession}".fasta | grep N50
done

@@

Hint: https://en.wikipedia.org/wiki/N50,_L50,_and_related_statistics ¨¨

@@

You will only perform the assembly on one genome, but use the loop form anyway. You will use the `--isolate` mode, but the `--careful` mode would also be interesting. This might take 10-15 minutes per genome.

@@

@@

Using Genome Tools, compare the result of the two assemblies

@@

@@

The goal here is to find ORFs and start codons, to find proteins in the genome.

@@

Find how prodigal works and use it to annotate the proteome of ERR029206.

```bash
prodigal -h
```

@@

@@

With the proteome of sample ERR029206, create a BLAST database. Use the `makeblastdb` function that comes with the BLAST package.

@@

```bash
module load
makeblastdb -dbtype prot -in ERR029206.faa -out ERR029206
```

@@

>>>>

The output redirection sign (`1>`) is different than the ususal `>`. It specifies that only the standard output (i.e. generally the results) will go to that file, while the standard error (i.e. warning and error messages, or other logging information) are displayed on the console. To save the standard error to a file, use `2>`, e.g.

```bash
skesa <options> 1> results.fasta 2> logfile
```

@@

@@

prints the help information in `skesa`

::::::::

@@

Find out how many contigs there are in the *M. tuberculosis* isolates. Use for example the [GenomeTools](https://genometools.org/) software, available as module with the same name on Uppmax. The main program is called `gt`. The documentation is comprehensive and can be found on the website or via the command line. We are looking for some subcommand that can give use statistics on sequences.

```bash
module load GenomeTools 
```
::::::::

```bash
gt --help
```
::::::::

```bash
gt seqstat --help
```
::::::::

@@

```bash
module load GenomeTools 
gt seqstat <fastafile>
```

@@

:::::::::::::::  instructor

@@

How to compare assemblies? What do N50 and L50 mean?

@@

Hint: https://en.wikipedia.org/wiki/N50,_L50,_and_related_statistics

Write loop over accessions, printing the accession name and running `gt` on each fasta file. Grepping for the information you want. You can grep two different patterns by prefixing each pattern with `-e`. E.g. `grep -e "this" -e "that"` greps this or that.

@@

```bash
for accession in <all accessions>
do
  echo <variable name, accessed with ${...}>
  gt seqstat <variable name>.fasta | grep -e <first pattern> -e <second patter>
done

@@

```bash
for accession in ERR029207 ERR029206 ERR026478 ERR026474 ERR026473 ERR026481 ERR026482
do
  echo "${accession}"
  gt seqstat "${accession}".fasta | grep -e N50 -e L50
done

@@

You will only perform the assembly on one genome, but use the loop form anyway. You will use the `--isolate` mode, but the `--careful` mode would also be interesting. This might take 5-10 minutes per genome.

SPAdes can be loaded with the `spades` module, and the main command is `spades.py`. Make a `spades_assembly` folder in the `molepi/results` folder and go there. Then run a 1-genome loop as above. SPAdes has arguments `-1` and `-2` for reads from two ends, the `-o` to use a specific folder to store the output (use `<accession>_spades`), `-t` to use several threads (use 2), and `-m` to set the maximum memory in Gb (use 8).

```bash

::::::::

@@

Folders etc:

```bash
cd /proj/g2020004/nobackup/3MK013/<username>/molepi/results
mkdir spades_assembly
cd spades_assembly
```

Running SPAdes

```bash
module load <module>
for accession in ERR029206
do
  <print the accession with echo>
  spades.py \
    <mode isolate> \
    <flag for read file 1> ../../data/trimmed_fastq/"${accession}"_1.fastq.gz_trim.fastq \
    <flag for read file 2> ../../data/trimmed_fastq/"${accession}"_2.fastq.gz_trim.fastq \
    -o <output folder> \
    <options for 2 threads and 8 Gb memory>
done

@@

Using Genome Tools, compare the result of the two assemblies.

```bash
cd /proj/g2020004/nobackup/3MK013/<username>/molepi/results
gt --help
```

:::

@@

```bash
cd /proj/g2020004/nobackup/3MK013/<username>/molepi/results
gt <sequence statistics?> <spades contigs>
gt <sequence statistics?> <skesa contigs>
```

::::::::

@@

The goal here is to find ORFs and start codons, and thus identify proteins in the genome.

@@

Find how prodigal works and use it to annotate the proteome of ERR029206. Make an `annotation` folder in the `molepi/results` folder. Use default settings, only setting the input file `-i` and writing the protein translations (`-a`) to a `.faa` file. Redirect the standard output (`>`) to a `.gbk` file. For both files, use the accession number as prefix.

```bash
module load prodigal
prodigal -h
```

:::

@@

```bash
cd /proj/g2020004/nobackup/3MK013/<username>/molepi/results
mkdir annotation
cd annotation
module load prodigal
prodigal -i <contig file> -a <accession.faa> > <accession.gbk>
```

@@

With the proteome (`.faa` file) of sample ERR029206, create a BLAST database, of protein type. Use the `makeblastdb` function that comes with the BLAST package.

```bash
module load blast
makeblastdb -h
```

@@

```bash
module load blast
makeblastdb <database type, protein> -in <proteome file> -out <accession>
```

```bash
module load blast
makeblastdb -dbtype prot -in ERR029206.faa -out ERR029206
```

@@

```bash
ls ../../../blast/
blastp -query <fasta file> -db accession
```
